---
title: "Getting Started with dissertationData"
author: "Catalina Canizares"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dissertationData}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Welcome to **dissertationData**! This package provides easy access to data from the **Youth Risk Behavior Survey (YRBS)**, a valuable resource for studying adolescent health trends.  

Whether you're an experienced analyst or just starting out, this guide will walk you through how to **load and work with the data in R** efficiently.

# Installation

If you haven't installed the package yet, you can install it from GitHub:

```{r eval=FALSE}
# Install from GitHub
install.packages("devtools")
devtools::install_github("yourusername/dissertationData")
```

Once installed, load the package:

```{r setup}
library(dissertationData)
```

# Working with the Data

This package provides different versions of the YRBS data. The simplest option is the raw datasets, available year by year from 2015 to 2023. 

You can access the data using two different functions: 
1. The base R function `data()`
2. A package-specific function `load_yrbs()`

Let's explore both approaches and their differences.

## Example: Examining Suicide Ideation in 2023 by Grade

To demonstrate how to work with the data, let’s analyze the percentage of suicide ideation in 2023 by grade.

### Option 1: Using the `data()` function

```{r}
data("raw2023")
```

This function loads the dataset into your environment as a **promise**, meaning that the object `raw2023` is only accessible once you interact with it through another function.

### Option 2: Using `load_yrbs()`

For more control over the object name and a direct approach to loading data, use the `load_yrbs()` function from this package:

```{r}
yrbs_df <- load_yrbs("raw2023")
```

This function immediately assigns the dataset to the `yrbs_df` object, making it easier to work with.

Let's continue interacting with the data to demonstrate how to answer questions using this dataset in the R environment.

I will proceed by selecting the variables I care most about, which are the grade and the suicide attempts items.

Considering we lost the data dictionaries due to recent federal actions, you can access the ones I was able to recover here. However, the package also includes dictionaries for each of the datasets.

Let's use one of the dictionaries.

```{r}
dict_2023 <- load_dictionary("2023")
dict_2023

```

Now that we have the dictionary loaded, let's search for the variables of interest: **suicide attempts** and **grade**. We can do this by querying the dictionary using `dplyr`.

```{r}
library(dplyr)
library(stringr)

# Filter for variables related to suicide attempts and grade
suicide_grade_vars <- dict_2023 %>%
  filter(str_detect(description, regex("suicide|grade", ignore_case = TRUE)))

# View the results
print(suicide_grade_vars)
```

This query will return all variables in the 2023 dataset that mention **suicide** or **grade** in their descriptions. This helps ensure we select the correct variables for our analysis.

Now that we have identified the variables of interest, **Q3** and **QN29** (all `QN` variables are dichotomized), let's create a summary table using these variables.


```{r}
library(gtsummary)
# Create a summary table for suicide attempts by grade

yrbs_df %>%
  select(Q3, QN29) %>%
  tbl_summary(
    by = QN29, 
    statistic = list(all_categorical() ~ "{p}%"),
    label = list(Q3 ~ "Grade")
  ) %>%
  modify_header(label ~ "Suicide Attempts") %>%
  modify_caption("**Table 1: Percentage of Suicide Attempts by Grade**")


```

Almost there! It would be great to clarify what "1" and "2" represent, as well as the grade levels corresponding to the options 1, 2, 3, 4, and 5.

Thanks to the dictionary, we know that 1 means "yes" and 2 means "no". This package provides a handy function, `binary_to_text()`, to easily convert these values into readable labels. Additionally, the `haven` package offers the `as_factor()` function, which can be used to convert categorical variables into factors, as demonstrated in the code below:


```{r}
yrbs_df %>%
  select(Q3, QN29) %>%
  mutate(QN29 = binary_to_text(QN29), 
         Q3 = haven::as_factor(Q3)) |> 
   tbl_summary(
    by = QN29, 
    statistic = list(all_categorical() ~ "{p}%"),
    label = list(Q3 ~ "Grade")
  ) %>%
  modify_header(label ~ "Suicide Attempts") %>%
  modify_caption("**Table 1: Percentage of Suicide Attempts by Grade**")
  
```


And just like that, we have a clear and well-formatted table showing that 9th graders have the highest proportion of students reporting suicidal thoughts.

Note: These proportions are calculated from the raw data and do not account for survey weights. If weighted estimates are needed, appropriate adjustments should be made before interpreting the results. We will cover how to properly apply survey weights in another vignette.

I hope this example is helpful in demonstrating how to use this package effectively within the R environment! 🚀


