---
title: "Getting Started with dissertationData"
author: "Catalina Canizares"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dissertationData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Welcome to **dissertationData**! This package provides easy access to data from the **Youth Risk Behavior Survey (YRBS)**, a valuable resource for studying adolescent health trends.  

Whether you're an experienced analyst or just starting out, this guide will walk you through how to **load and work with the data in R** efficiently.

# Installation

If you haven't installed the package yet, you can install it from GitHub:

```{r eval=FALSE}
# Install from GitHub
install.packages("devtools")
devtools::install_github("yourusername/dissertationData")
```

Once installed, load the package:

```{r setup}
library(dissertationData)
```

# Working with the Data

This package provides different versions of the YRBS data. The simplest option is the raw datasets, available year by year from 2015 to 2023. 

You can access the data using two different functions: 
1. The base R function `data()`
2. A package-specific function `load_yrbs()`

Let's explore both approaches and their differences.

## Example: Examining Suicide Ideation in 2023 by Grade

To demonstrate how to work with the data, let’s analyze the percentage of suicide ideation in 2023 by grade.

### Option 1: Using the `data()` function

```{r}
data("raw2023")
```

This function loads the dataset into your environment as a **promise**, meaning that the object `raw2023` is only accessible once you interact with it through another function.

### Option 2: Using `load_yrbs()`

For more control over the object name and a direct approach to loading data, use the `load_yrbs()` function from this package:

```{r}
yrbs_df <- load_yrbs("raw2023")
```

This function immediately assigns the dataset to the `yrbs_df` object, making it easier to work with.

Let's continue interacting with the data to demonstrate how to answer questions using this dataset in the R environment.

I will proceed by selecting the variables I care most about, which are the grade and the suicide attempts items.

Considering we lost the data dictionaries due to recent federal actions, you can access the ones I was able to recover here. However, the package also includes dictionaries for each of the datasets.

Let's use one of the dictionaries.

