---
title: "Saving YRBS Data to Your Computer"
author: "Francisco Cardozo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Saving YRBS Data to Your Computer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Welcome to **dissertationData**! This package provides easy access to data from the **Youth Risk Behavior Survey (YRBS)**, a critical resource for studying adolescent health trends.  

Whether you're an experienced analyst or just starting, this guide will walk you through how to **load, and save YRBS data** efficiently.  

# Installation

If you haven't installed the package yet, you can install it from GitHub:

```{r eval=FALSE}
# Install from GitHub:
# install.packages("devtools")
# devtools::install_github("yourusername/dissertationData")
```

Once installed, load the package:

```{r setup}
library(dissertationData)
```

# Saving a Local Copy of the Data File

There are times when you might want to store a local copy of the raw data for your own use. This package allows you to save the data in multiple formats, including CSV, SPSS (.dat), and Parquet, so you can choose the format that works best for you.

Many users prefer to save data in **CSV format**. This package provides a dedicated function that allows you to save the dataset as a CSV file while automatically creating a folder called `"data"` to store it.  

This function allows you to save any dataset available in the package as a CSV file.  To check the available datasets, use:  

```{r}
list_package_datasets("dissertationData")
```

To save the raw 2015 dataset as a CSV file, use the following example: 

```{r eval=FALSE}
# Save the data in a csv form
save_data_to_csv("raw2015")  
```

After running this command, the dataset will be saved in your local machine, and the file path will be displayed in the console. For example:

`Dataset saved successfully at: /Users/myuser/myfolder/data/raw2015.csv`

# Downloading Raw YRBS Data in Different Formats
If you only need the **raw version** of the YRBS dataset in a format other than CSV, we have designed a function that gives you full control over:  

- **The year** of the dataset you need  
- **The folder** where you want to store the file  
- **The file format**, choosing from **CSV, Parquet, SPSS (.sav), or RDS**  

The function is called `get_raw_data()`, and you can specify the storage folder using the `dest_dir` parameter.  

For example, to save the **2017 dataset** in a folder named `"my_data"`, run:  

```{r eval=FALSE}
# Copy the YRBS 2017 SPSS file to a folder called "my_data"
save_raw_data(2017, format = "spss", dest_dir = "my_data")
```

If the specified folder does not exist, the function will **ask for your permission** to create it.  
Once confirmed, it will generate the folder and provide the file's location.  

For example, if you specify `"my_data"` as the storage folder but it doesn’t exist, you'll see:  

`Would you like to proceed with creating '{dest_dir}'? (y/n): y`
`✔ Directory 'my_data' created.`
`✔ File copied to my_data/yrbs2017.sav`

# Summary

In this vignette we covered:

* Installing and loading the package
* Saving a copy of the raw data file to your computer
* Choosing different data formats
* Understanding the error messages when an invalid year is used

The goal of dissertationData is to simplify the process of accessing and working with YRBS data. We hope this guide helps you get started quickly and confidently.

If you have any questions or run into issues, please refer to the package documentation or contact the package maintainer.

Happy analyzing!
